---
title: 'DATA 605 - Homework #10'
author: "Zach Alexander"
date: "4/4/2020"
output: html_document
---

***

> **Question**: Smith is in jail and has 1 dollar; he can get out on bail if he has 8 dollars. A guard agrees to make a series of bets with him. If Smith bets A dollars, he wins A dollars with probability .4 and loses A dollars with probability .6.

> Find the probability that he wins 8 dollars before losing all of his money if:
(a) he bets 1 dollar each time (timid strategy).

In order to solve for this problem, we can utilize the Gambler's Ruin equation from out textbook: 

$P = \frac{1 - (\frac{q}{p})^{s}}{1 - (\frac{q}{p})^{M}}$  

where $M = 8$, $s = 1$, $p = 0.4$, and $q = 0.6$, since the probability of Smith winning A dollars is 0.4, losing A dollars is 0.6, Smith starts with a stake of 1 (1 dollar), and it'll be played until Smith reaches a capital of 8.  

With this in mind, we can solve the probability:  

$P = \frac{1 - (\frac{0.6}{0.4})^{1}}{1 - (\frac{0.6}{0.4})^{8}}$  

```{r}
prob <- (1 - (0.6/0.4)^1) / (1 - (0.6/0.4)^8)
prob
```

**ANSWER: The probability that Smith wins 8 dollars before losing all of his money if he bets 1 dollar each time (and applies a timid strategy) is about 0.0203.**  

We can also solve this problem using matrices and Markov chains. First, we can set up our transition matrix  

```{r}
P <- matrix(c(0,1,0,0,0,0,0,0,0,
              0.6,0,0.4,0,0,0,0,0,0,
              0,0.6,0,0.4,0,0,0,0,0,
              0,0,0.6,0,0.4,0,0,0,0,
              0,0,0,0.6,0,0.4,0,0,0,
              0,0,0,0,0.6,0,0.4,0,0,
              0,0,0,0,0,0.6,0,0.4,0,
              0,0,0,0,0,0,0.6,0,0.4,
              0,0,0,0,0,0,0,0,1), nrow=9, byrow=TRUE)
rownames(P) <- c("0","1","2","3","4","5","6","7","8")
colnames(P) <- c("0","1","2","3","4","5","6","7","8")
P

# Matrix with just transient states
Q <- P[2:8, c(2:8)]
Q
# Matrix with just absorbing states
R <- P[2:8, c(1,9)]
R
```

Now, with the matrices separated into transient and absorbing states, we can use matrix decomposition to break down the matrix with transient states ($N = (I - Q)^{-1}$): 
```{r}
# Fundamental matrix
I <- diag(7)
N <- solve(I-Q)
N

```

Finally, we can find our absorption probabilities:
```{r}
# Absorption probablities
X <- N %*% R
X
```

As we can see, if we just isolate $s = 1$, we get:  

```{r}
paste0('The probability starting with a stake of 1 is ', round(X["1","8"], 4), '.')
```
***

> (b) he bets, each time, as much as possible but not more than necessary to bring his fortune up to 8 dollars (bold strategy).

Now, if Smith takes a bolder approach in his betting strategy, the transition matrix will look like:  

\begin{equation}
  P = 
  \begin{bmatrix}
  1 & 0 & 0 & 0 & 0\\ 
  0.6 & 0 & 0.4 & 0 & 0\\ 
  0.6 & 0 & 0 & 0.4 & 0\\ 
  0.6 & 0 & 0 & 0 & 0.4\\ 
  0 & 0 & 0 & 0 & 1\\ 
  \end{bmatrix}
\end{equation}

With the columns of the matrix indicating transitions between 0, 1, 2, 4, and 8 dollars:  
```{r}
P2 <- matrix(c(1,0,0,0,0,
               0.6,0,0.4,0,0,
               0.6,0,0,0.4,0,
               0.6,0,0,0,0.4,
               0,0,0,0,1), ncol=5,nrow=5, byrow = TRUE)
rownames(P2) <- c("1","2","3","4","5")
colnames(P2) <- c("0","1","2","4","8")
P2
```

Since our initial state, matching our transition matrix above, will start with 1 dollar:

\begin{equation}
  initial = 
  \begin{bmatrix}
  0 & 1 & 0 & 0 & 0\\ 
  \end{bmatrix}
\end{equation}  


```{r}
initial <- matrix(c(0, 1, 0, 0, 0), ncol=5,nrow = 1,byrow = TRUE)
```

With the transition matrix set, and the inital state ready as well, we can apply this to our Markov chain:  

```{r}
p1 <- initial %*% P2
p1
```
```{r}
p2 <- p1 %*% P2
p2
```

```{r}
p3 <- p2 %*% P2
p3
```

```{r}
p4 <- p3 %*% P2
p4
```

**ANSWER: We can see that if Smith takes a bolder approach and bets each time, as much as possible but not more than necessary to bring his fortune up to 8 dollars, he'll have about a 0.064 probability of getting out on bail.**  

***

> (c) Which strategy gives Smith the better chance of getting out of jail?

**ANSWER: As we can see from the calculation of our probabilites in parts a and b, it would be wiser for Smith to take the bolder approach in order to have a better chance of getting out on bail and obtaining the 8 dollars from betting.**  
