---
title: "DATA 605 - Homework 2"
author: "Zach Alexander"
date: "2/8/2020"
output: html_document
---


***

### PROBLEM SET #1

***

> Question #1: Show that $A^{T}A \neq AA^{T}$ in general. (Proof and demonstration.) 

***

For our proof of $A^{T}A \neq AA^{T}$, we assume that A is not a square matrix, or $m \neq n$.  

Therefore, where $A$ has dimensions of an $m \times n$ matrix and $A^{T}$ has dimensions of an $n \times m$ matrix:  

The product of $A^{T}A$ will be an $n \times n$ matrix and the product of $AA^{T}$ will be an $m \times m$ matrix. Since we know that $m \neq n$, then $A^{T}A \neq AA^{T}$. This will be true for all non-square matrix.  

**Demonstration:**  

We have a $3 \times 2$ matrix, $A$:

```{r}
A <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 3, ncol = 2)
A
```

Therefore, $A^{T}$ is:  

```{r}
At <- t(A)
At
```

When we test to see whether the equation of $A^{T}A \neq AA^{T}$ is true, we find the following:

```{r}
prodAAt <- A%*%At
prodAAt
```

Where `prodAAt` is a $3 \times 3$ matrix. And:

```{r}
prodAtA <- At%*%A
prodAtA
```

Where `prodAtA` is a $2 \times 2$ matrix. We can see that `prodAAt` and `prodAtA` are not equal.

***

> Question #2: For a special type of square matrix $A$, we get $A^{T}A = AA^{T}$. Under what conditions could this be true? (Hint: The Identity matrix I is an example of such a matrix).

***

For a special type of square matrix, where $A = A^{T}$, this condition could be true. Taking the Identity matrix, $I$ as an example:

\begin{bmatrix}
1 & 0 & 0\\ 
0 & 1 & 0\\ 
0 & 0 & 1
\end{bmatrix}

For $I$ above, we can see that $I^{T}$ would be the same as $I$. This also is true for other square matrices where $A = A^{T}$.  

**Demonstration:**  

```{r}
mat1 <- matrix(c(1, 0, 0, 0, 1, 0, 0, 0, 1), nrow = 3, ncol = 3)
mat1

mat1T <- t(mat1)
mat1T
```

Another example is:  

```{r}
mat2 <- matrix(c(1, 6, -10, 6, 12, 5, -10, 5, 1), nrow = 3, ncol = 3)
mat2

mat2T <- t(mat2)
mat2T
```

Therefore, we can prove from these examples that these special square matrices make $A^{T}A = AA^{T}$ true:

```{r}
mat1T%*%mat1 == mat1%*%mat1T
mat2T%*%mat2 == mat2%*%mat2T
```

***

### PROBLEM SET #2

***

> Write an R function to factorize a square matrix $A$ into $LU$ or $LDU$, whichever you prefer.

***

In order to work through this problem, we need to keep in mind that $A = LU$. Therefore, we can write our function as follows:

Since I developed similar code to this last week to solve for row echelon form (and created an Upper Triangular Matrix), I'll adapt this to start my a `factorizeInitial()` function:

```{r}
factorizeInitial <- function(a){
  u <- a
  l <- matrix(c(1, 0, 0, 0, 1, 0, 0, 0, 1), nrow = 3, ncol = 3)
  
  if(a[2, 1] != 0) {
    mult_1 <- -(u[2, 1] / u[1, 1])

    u[2,] <- (u[1,] * mult_1) + u[2,]
    l[2, 1] <- -mult_1
  }

  if(a[3, 1] != 0) {
    mult_2 <- -(u[3, 1] / u[1, 1])

    u[3,] <- (u[1,] * mult_2) + u[3,]
    l[3, 1] <- -mult_2
  }

  if(a[3, 2] != 0) {
    mult_3 <- -(u[3, 2] / u[2, 2])

    u[3,] <- (u[2,] * mult_3) + u[3,]
    l[3, 2] <- -mult_3
  }
  print(a)
  print(l)
  print(u)
}
```

Initial test:

```{r}
mat3 <- matrix(c(2, 1, -6, 4, -4, -9, -4, 3, 5), nrow = 3, ncol = 3)

factorizeInitial(mat3)
```

Great! It looks like I have it working for a $3 \times 3$ matrix. However, the problem would like us to account for different sized square matrices. I'm also seeing patterns within my initial function, so I did my best to make the code more efficient. In the function `factorizeM()`, I was able to do so below:

```{r}
factorizeM <- function(a){
  u <- a
  m <- dim(a)[1]
  n <- dim(a)[2]
  l <- diag(m)
  
  # We need to first check that a is a square matrix
  if (m != n) {
    return('This is a non-square matrix.')
  }
  
  # We then need to check if it's a one by one matrix
  if (m==1 & n==1) {
    return(list(a, l, u, a == l%*%u))
  }
  
  for(i in 2:m){
    for(j in 1:(i-1)){
      mult <- -(u[i, j] / u[j, j])
      u[i,] <- (u[j,] * mult) + u[i,]
      l[i, j] <- -mult
    }
  }
  print(a)
  print(u)
  print(l)
  
  # check to ensure it is correct
  a == l%*%u
}
```

Let's test our `factorizeM()` function to make sure we get the same matrices as `factorizeInitial()`:  

```{r}
mat3 <- matrix(c(2, 1, -6, 4, -4, -9, -4, 3, 5), nrow = 3, ncol = 3)

factorizeM(mat3)
```

We can see that this does occur! Additionally, I added a check into the `factorizeM()` function to check our final matrices to prove $A = LU$.

I've also performed tests on this function for a non-square matrix, a $1 \times 1$, $2 \times 2$, and $4 \times 4$ matrix below to ensure it still works correctly:

**Non-square matrix example**

```{r}
nonsquare_mtx <- matrix(c(3, 2, 3, 1, 2, 5), nrow = 2, ncol = 3)
factorizeM(nonsquare_mtx)
```

**$1 \times 1$ matrix example**
```{r}
onebyone_mtx <- matrix(c(6), nrow = 1, ncol = 1)
factorizeM(onebyone_mtx)
```


**$2 \times 2$ matrix example**
```{r}
twobytwo_mtx <- matrix(c(2, 3, 4, 2), nrow = 2, ncol = 2)
factorizeM(twobytwo_mtx)
```

**$4 \times 4$ matrix example**
```{r}
fourbyfour_mtx <- matrix(c(2, 3, 4, 2, 8, 2, 3, 0, 3, 5, 7, 3, 4, 3, 6, 7), nrow = 4, ncol = 4)
factorizeM(fourbyfour_mtx)
```
